Multi-resolution Neural Field Estimation

### 2011-03-03 Multiresolution_framework/ltx/manuscript.tex
a few minor edits

\section{IDE Neural Field Model}
The stochastic integro-difference equation (IDE) form of the Amari neural field  formulation~\cite{Amari1977} is given by (see~\cite{Freestone2011} for a full derivation)
\begin{equation}\label{eq:DiscreteTimeModel}
	v_{t+1}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r'}\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ is denoted $v_t\left(\mathbf r\right)$. Synaptic dynamics are included through the parameter $\xi=\left(1-\frac{ T_s}{\tau}\right)$, where $\tau$ is the synaptic time constant and $T_s$ is the sampling time. The connectivity strength between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$. The connectivity kernel is taken as a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977}. The term $e_t(\mathbf r)$ is an $i.i.d.$ disturbance such that $e_t(\mathbf{s})\sim\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$, where $\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$  denotes a zero mean spatial Gaussian process with covariance function $\eta(\mathbf{r}-\mathbf{r'})$~\cite{Rasmussen2005}. The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the activation function \dean{$f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r}) + v_0$}. A linear form of the activation function was chosen to construct the estimator due to high computational demands when using a nonlinear (sigmoidal) activation function. The observation equation describing the electrophysiological recordings is given by 
\begin{equation}\label{eq:ObservationEquation}
	\mathbf y_t = \int_{\Omega} { m\left(\mathbf{r}_{n}-\mathbf{r}'\right) v_t\left(\mathbf{r}'\right) \, d\mathbf{r}'} + \boldsymbol\epsilon_t, 
\end{equation}
where $m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right)$ is the observation kernel and the observation noise is an i.i.d. white noise process and $\boldsymbol{\epsilon}_{t}\sim \mathcal{N}\left(\mathbf{0},\mathbf{\Sigma}_{\epsilon}\right)$. % $\mathbf{y}_{t} = [y_t(\mathbf{r}_1) y_t(\mathbf{r}_2) \cdots y_t(\mathbf{r}_{n_y})]^\top$, compiled at $n_{y}$ spatial locations at time $t$ via the observation kernel , is corrupted by The superscript $\top$ denotes the transpose operator.
\section{MRA of the IDE in State-Space}
The multi-resolution approximation (MRA) of the neural IDE is obtained by decomposing both the field, $v_t(.)$, and the connectivity kernel, $w(.)$, (assuming square-integrable functions) using translations and dilations of a scaling function $\phi(r)$ and a mother wavelet $\psi(r)$. Considering a one-dimensional field, the connectivity kernel is decomposed by,
\begin{equation}
 w\left(r-r'\right)=\sum_{l \in \mathbb{Z}}\alpha_{j_0,l}\phi_{j_0,l}\left(r-r'\right)+\sum_{j\geq j_0}^{\infty} \sum_{l \in \mathbb{Z}}\beta_{j,l}\psi_{j,l}\left(r-r'\right), 
\label{eq:KernelExpansion}
\end{equation}
where $\alpha_{j_0,l}$ are the approximation coefficients at the lowest scale $j_0$, and $\beta_{j,l}$ are the detail coefficients at different scales $j$, with $\phi_{j,l}\left(r\right)=2^{\frac{j}{2}}\phi\left(2^jr-l\right) $ and $\psi_{j,l}\left(r\right)=2^{\frac{j}{2}}\psi\left(2^jr-l\right)$. Integers $j$ and $l$ are the scale and translation parameters respectively. The field is decomposed as
\begin{equation}
 v_t\left(r\right)=\sum_{l \in \mathbb{Z}}x_{t,j_{0},l}\phi_{j_{0},l}\left(r\right)+\sum_{j\geq j_0}^{\infty} \sum_{l \in \mathbb{Z}} \check{x}_{t,j,l}\psi_{j,l}\left(r\right),
\label{eq:FieldExpansion}
\end{equation}
where $x_{t,j_{0},l}$ and $\check{x}_{t,j,l} $ are the coefficients of the expansion at time $t$. \dean{Need to define these guys as the states here!}

@@

 \boldsymbol\upsilon_0 & =\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\odot\mathbf U]_{j,i}\\ 
 \boldsymbol\upsilon_1 & =\xi\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\odot\mathbf U]_{j,i}\label{eq:upsilon1} \\

@@

where $[.]_{j:} $ denotes $j^{th}$ row of the matrix. The matrices $\Xi_0$ and $\Xi_1$ are calculated using the Rauch Tung Streibel smoother \cite{RAUCH1965} outputs: state estimates, $\hat{\mathbf x}_t$, covariance, $\mathbf P_t=\mathrm{cov}(\mathbf{x_t})$, and cross-covariance matrix $\mathbf M_t=\mathrm{cov}(\mathbf{x}_{t-1},\mathbf{x}_{t})$ \cite{Gibsona2005}

@@

\section{Results}
To demonstrate the performance of the MRA estimation framework, data was generated synthetically  using Eqs.~\ref{eq:DiscreteTimeModel} and \ref{eq:ObservationEquation}, enabling a comparison between true and estimated parameters. 100 realizations of 1 second of data were generated where $\boldsymbol\Sigma_v=0.1 \times \mathbf{I}_{n_y}$ \dean{and the disturbance covariance was a B-spline with $j=3$.} 
The sampling period and the synaptic time constant were set to $T_s = 1$~ms and $\tau = 10$~ms, respectively. The distance between adjacent sensors was $0.5$~mm resulting into 161 observations. \parham{Since $n_x<161$ at each level, this provides enough information for state estimation due to orthogonality of the basis functions.}  The observation kernel is modeled using a B-spline function, with a width of 0.8~mm at half the maximum amplitude of the observation kernel. The spacing and bandwidth of the sensors allowed for the full spatial bandwidth of the field to be observed. The spatial cutoff frequency of the observed field, $y_t(r_n)$, was used to specify the level of decomposition required to represent the field using the wavelets. The spatial cutoff frequency was \dean{??} cycles/mm. Therefore, wavelets up to level $j=2$ (with the bandwidth $\approx[2.5,4]$ cycles/mm) can represent the significant spatial characteristics of the field. \cut{Note that wavelets due to their bandpass filter characteristics extract successively higher and higher frequency components by increasing $j$.}

>>>>

\section{IDE Neural Field Model}
The stochastic integro-difference equation (IDE) form of the Amari neural field  formulation~\cite{Amari1977} is given by (see~\cite{Freestone2011} for a full derivation)
\begin{equation}\label{eq:DiscreteTimeModel}
	v_{t+1}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r'}\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ is denoted $v_t\left(\mathbf r\right)$. Synaptic dynamics are included through the parameter $\xi=\left(1-\frac{ T_s}{\tau}\right)$, where $\tau$ is the synaptic time constant and $T_s$ is the sampling time. The connectivity strength between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$. The connectivity kernel is taken as a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977}. The term $e_t(\mathbf r)$ is an i.i.d. disturbance such that $e_t(\mathbf{s})\sim\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$, where $\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$  denotes a zero mean spatial Gaussian process with covariance function $\eta(\mathbf{r}-\mathbf{r'})$~\cite{Rasmussen2005}. The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the activation function \dean{$f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r}) + v_0$}. A linear form of the activation function was chosen to construct the estimator due to high computational demands when using a nonlinear (sigmoidal) activation function. The observation equation describing the electrophysiological recordings is given by 
\begin{equation}\label{eq:ObservationEquation}
	\mathbf y_t = \int_{\Omega} { m\left(\mathbf{r}_{n}-\mathbf{r}'\right) v_t\left(\mathbf{r}'\right) \, d\mathbf{r}'} + \boldsymbol\epsilon_t, 
\end{equation}
where $m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right)$ is the observation kernel and the observation noise is an i.i.d. white noise process and $\boldsymbol{\epsilon}_{t}\sim \mathcal{N}\left(\mathbf{0},\mathbf{\Sigma}_{\epsilon}\right)$. % $\mathbf{y}_{t} = [y_t(\mathbf{r}_1) y_t(\mathbf{r}_2) \cdots y_t(\mathbf{r}_{n_y})]^\top$, compiled at $n_{y}$ spatial locations at time $t$ via the observation kernel , is corrupted by The superscript $\top$ denotes the transpose operator.
\section{MRA of the IDE in State-Space}
The multi-resolution approximation (MRA) of the neural IDE is obtained by decomposing both the field, $v_t(.)$, and the connectivity kernel, $w(.)$, (assuming square-integrable functions) using translations and dilations of a scaling function $\phi(r)$ and a mother wavelet $\psi(r)$. Considering a one-dimensional field, the connectivity kernel is decomposed by,
\begin{equation}
 w\left(r-r'\right)=\sum_{l \in \mathbb{Z}}\alpha_{j_0,l}\phi_{j_0,l}\left(r-r'\right)+\sum_{j\geq j_0}^{\infty} \sum_{l \in \mathbb{Z}}\beta_{j,l}\psi_{j,l}\left(r-r'\right), 
\label{eq:KernelExpansion}
\end{equation}
where $\alpha_{j_0,l}$ are the approximation coefficients at the lowest scale $j_0$, and $\beta_{j,l}$ are the detail coefficients at different scales $j$, with $\phi_{j,l}\left(r\right)=2^{\frac{j}{2}}\phi\left(2^jr-l\right) $ and $\psi_{j,l}\left(r\right)=2^{\frac{j}{2}}\psi\left(2^jr-l\right)$. Integers $j$ and $l$ are the scale and translation parameters respectively. The field is decomposed as
\begin{equation}
 v_t\left(r\right)=\sum_{l \in \mathbb{Z}}x_{t,j_{0},l}\phi_{j_{0},l}\left(r\right)+\sum_{j\geq j_0}^{\infty} \sum_{l \in \mathbb{Z}} \check{x}_{t,j,l}\psi_{j,l}\left(r\right),
\label{eq:FieldExpansion}
\end{equation}
where $x_{t,j_{0},l}$ and $\check{x}_{t,j,l} $ are the coefficients of the expansion at time $t$. \dean{Need to define these guys as the states here!}

@@

 \boldsymbol\upsilon_0 & =\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}[\boldsymbol\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\odot\mathbf U]_{j,i}\\ 
 \boldsymbol\upsilon_1 & =\xi\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\boldsymbol\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\odot\mathbf U]_{j,i}\label{eq:upsilon1} \\

@@

where $[.]_{j:} $ denotes $j^{th}$ row of the matrix. The matrices $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ are calculated using the Rauch Tung Streibel smoother \cite{RAUCH1965} outputs: state estimates, $\hat{\mathbf x}_t$, covariance, $\mathbf P_t=\mathrm{cov}(\mathbf{x_t})$, and cross-covariance matrix $\mathbf M_t=\mathrm{cov}(\mathbf{x}_{t-1},\mathbf{x}_{t})$ \cite{Gibsona2005}

@@

\section{Results}
To demonstrate the performance of the MRA estimation framework, data was generated synthetically  using Eqs.~\ref{eq:DiscreteTimeModel} and \ref{eq:ObservationEquation}, enabling a comparison between true and estimated parameters. 100 realizations of 1 second of data were generated where $\boldsymbol\Sigma_v=0.1 \times \mathbf{I}_{n_y}$ and $\eta(r-r')=\phi_{3,-2}(r-r')$ (\dean{Defining the covariance this way is confusing, just like it was for the sensor kernel. It implies that the disturbance covariance to directly related to the field, or the basis was chosen because of the disturbance. Can we simply say that the disturbance covariance was a b-spline with dilation xx?}). 
The sampling period and the synaptic time constant were set to $T_s = 1$~ms and $\tau = 10$~ms, respectively. The distance between adjacent sensors was $0.5$~mm resulting into 161 observations. \parham{Since $n_x<161$ at each level, this provides enough information for state estimation due to orthogonality of the basis functions.}  The observation kernel is modeled using a B-spline function, with a width of 0.8~mm at half the maximum amplitude of the observation kernel. The spacing and bandwidth of the sensors allowed for the full spatial bandwidth of the field to be observed. The spatial cutoff frequency of the observed field, $y_t(r_n)$, was used to specify the level of decomposition required to represent the field using the wavelets. The spatial cutoff frequency was \dean{??} cycles/mm. Therefore, wavelets up to level $j=2$ (with the bandwidth $\approx[2.5,4]$ cycles/mm) can represent the significant spatial characteristics of the field. \cut{Note that wavelets due to their bandpass filter characteristics extract successively higher and higher frequency components by increasing $j$.}
